---
title: 'bioinf545 project'
author: "<h3>Nikki Watson</h3>"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: spacelab
    highlight: tango
    toc: yes
    number_sections: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
    self_contained: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
tags:
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r update}
# R.version.string
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# # Install clusterProfiler
# BiocManager::install("clusterProfiler")
# BiocManager::install("org.Hs.eg.db")
# BiocManager::install("GO.db")
# BiocManager::install("SingleR")
# BiocManager::install("celldex")
```


```{r load packages}
library(ggplot2)
library(Seurat)
library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(GO.db)
library(celldex)
library(SingleR)
library(tidyr)
library(knitr)
library(kableExtra)
```

3 = control
4 = disease

```{r library prep}
# Pool 3
data_dir3 <- "C:/Users/User/University of Michigan Dropbox/Nikki Watson/HP Backup/umich/courses/bioinf545/project/Pool3_ad"
adata3 <- Read10X(data.dir = data_dir3)
seurat_object3 <- CreateSeuratObject(counts = adata3[["Gene Expression"]]) 

# Pool 4
data_dir4 <- "C:/Users/User/University of Michigan Dropbox/Nikki Watson/HP Backup/umich/courses/bioinf545/project/Pool4_ad"
adata4 <- Read10X(data.dir = data_dir4)
seurat_object4 <- CreateSeuratObject(counts = adata4[["Gene Expression"]])

```


```{r count}
calculate_cell_metrics <- function(seurat_object) {
  # Use the correct layer, probably "counts", or check your specific Seurat object structure
  counts_matrix <- GetAssayData(seurat_object, layer = "counts")
  
  # Calculate total counts per cell
  seurat_object[["nCounts"]] <- Matrix::colSums(counts_matrix)
  
  # Calculate number of genes with counts per cell
  seurat_object[["nGenes"]] <- Matrix::colSums(counts_matrix > 0)
  
  return(seurat_object)
}

# Pool 3
seurat_object3 <- calculate_cell_metrics(seurat_object3)
seurat_object4 <- calculate_cell_metrics(seurat_object4)
```

```{r QC}
# Pool 3
VlnPlot(seurat_object4, features = c("nCounts", "nGenes"), pt.size = 0.4)

# Pool 4
VlnPlot(seurat_object3, features = c("nCounts", "nGenes"), pt.size = 0.4)

```

```{r scatter plots}
# Pool 3
FeatureScatter(seurat_object3, feature1 = "nCounts", feature2 = "nGenes")

# Pool 4
FeatureScatter(seurat_object4, feature1 = "nCounts", feature2 = "nGenes")

```

```{r filter}
filter_seurat_object <- function(seurat_object) {
  # Filter cells with more than 500 counts
  seurat_object <- subset(seurat_object, subset = nCounts > 500)

  # Retrieve the correct data layer for counts
  counts_data <- GetAssayData(seurat_object, layer = "counts")  # Adjust 'counts' to your actual data layer name if needed

  # Check if retrieval is successful:
  if(!is.null(counts_data)) {
    # Filter genes present in at least 3 cells
    expressed_genes <- rowSums(counts_data > 0) > 3
    # Subset the Seurat object only to include those genes
    seurat_object <- subset(seurat_object, features = names(which(expressed_genes)))
  } else {
    stop("Counts data could not be retrieved. Please check the data layer name.")
  }
  
  return(seurat_object)
}

# Pool 3
seurat_object3 <- filter_seurat_object(seurat_object3)
seurat_object4 <- filter_seurat_object(seurat_object4)

```

```{r normalize}
# Pool 3
seurat_object3 <- NormalizeData(seurat_object3, normalization.method = "LogNormalize", scale.factor = 1e4)

# Pool 4
seurat_object4 <- NormalizeData(seurat_object4, normalization.method = "LogNormalize", scale.factor = 1e4)

```

```{r variable genes}
# Pool 3
seurat_object3 <- FindVariableFeatures(seurat_object3, selection.method = "vst", nfeatures = 2000)

# Pool 4
seurat_object4 <- FindVariableFeatures(seurat_object4, selection.method = "vst", nfeatures = 2000)

```

```{r plot variable}
# Pool 3
VariableFeaturePlot(seurat_object3)

# Pool 4
VariableFeaturePlot(seurat_object4)
```

```{r store raw}
# Pool 3
rna_counts3 <- GetAssayData(seurat_object3, layer = "counts") 
seurat_object3 <- SetAssayData(seurat_object3, layer = "RNA_counts", new.data = rna_counts3)

# Pool 4
rna_counts4 <- GetAssayData(seurat_object4, layer = "counts")
seurat_object4 <- SetAssayData(seurat_object4, layer = "RNA_counts", new.data = rna_counts4)

```

```{r scale}
# Pool 3
seurat_object3 <- ScaleData(seurat_object3, features = rownames(seurat_object3), maxChunkSize = 10)

# Pool 4
seurat_object4 <- ScaleData(seurat_object4, features = rownames(seurat_object4), maxChunkSize = 10)
```

```{r PCA}
# Pool 3
seurat_object3 <- RunPCA(seurat_object3, npcs = 50)

# Pool 4
seurat_object4 <- RunPCA(seurat_object4, npcs = 50)
```

```{r plot PCA}
# Pool 3
ElbowPlot(seurat_object3, ndims = 50)

# Pool 4
ElbowPlot(seurat_object4, ndims = 50)
```

```{r UMAP}
# Pool 3
seurat_object3 <- FindNeighbors(seurat_object3, dims = 1:40)
seurat_object3 <- RunUMAP(seurat_object3, dims = 1:40)

# Pool 4
seurat_object4 <- FindNeighbors(seurat_object4, dims = 1:40)
seurat_object4 <- RunUMAP(seurat_object4, dims = 1:40)
```

```{r clustering}
# Pool 3
seurat_object3 <- FindClusters(seurat_object3, resolution = 0.5)

# Pool 4
seurat_object4 <- FindClusters(seurat_object4, resolution = 0.5)
```

```{r plot UMAP}
# Pool 3
DimPlot(seurat_object3, reduction = "umap", group.by = "seurat_clusters", label = TRUE) +
  labs(title = "UMAP Clusters - Pool 3 (Control)")
FeaturePlot(seurat_object3, features = c("nCounts", "nGenes"))

# Pool 4
DimPlot(seurat_object4, reduction = "umap", group.by = "seurat_clusters", label = TRUE) +
  labs(title = "UMAP Clusters - Pool 4 (Disease)")
FeaturePlot(seurat_object4, features = c("nCounts", "nGenes"))
```

```{r marker genes}
# Pool 3
markers3 <- FindMarkers(seurat_object3, ident.1 = "0", test.use = "t")

# Pool 4
markers4 <- FindMarkers(seurat_object4, ident.1 = "0", test.use = "t")
```

```{r top markers}
# Pool 3
DoHeatmap(seurat_object3, features = head(rownames(seurat_object3), 20))

# Pool 4
DoHeatmap(seurat_object4, features = head(rownames(seurat_object4), 20))
```


Score cells for cell cycle phases. Assigns a score to each cell indicating its position in the cell cycle phases (G1, S, G2/M) based on gene expression patterns

Used to determine if cell cycle effects influence data
```{r cell cycle}
s.genes <- Seurat::cc.genes$s.genes
g2m.genes <- Seurat::cc.genes$g2m.genes

# Pool 3
seurat_object3 <- CellCycleScoring(seurat_object3, s.features = s.genes, g2m.features = g2m.genes)

# Pool 4
seurat_object4 <- CellCycleScoring(seurat_object4, s.features = s.genes, g2m.features = g2m.genes)

# Visualize scores
VlnPlot(seurat_object3, features = c("S.Score", "G2M.Score"))
VlnPlot(seurat_object4, features = c("S.Score", "G2M.Score"))
```


Aligning cells across dataset to minimize batch effects
```{r data integration}
seurat_object_list <- list(pool3 = seurat_object3, pool4 = seurat_object4)

# Identify integration anchors
anchors <- FindIntegrationAnchors(object.list = seurat_object_list)

# Integrate data
integrated <- IntegrateData(anchorset = anchors)

# Use integrated data for downstream analysis
integrated <- ScaleData(integrated)
integrated <- RunPCA(integrated)
integrated <- RunUMAP(integrated, dims = 1:20)
DimPlot(integrated, reduction = "umap", group.by = "orig.ident")
```

Identify genes that are differentially expressed between cluster
Used for discovering markers of specific cell types or identifying genes linked to disease states
```{r differential expression}
#Pool 3
markers3 <- FindAllMarkers(seurat_object3, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
markers3[1:20,]

#Pool 4
markers4 <- FindAllMarkers(seurat_object4, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
markers4[1:20,]
```


Assess which biological pathways are enriched among the differentially expressed genes, provide insights into function
Identify key pathways involved
```{r pathway analysis}
# Pool 3
gene_names <- markers3$gene
log2fc_values <- markers3$avg_log2FC

## convert gene symbols to ENTREZ IDs
entrez_ids <- mapIds(
  org.Hs.eg.db,
  keys = gene_names,
  column = "ENTREZID",
  keytype = "SYMBOL",
  multiVals = "first"  # getting the first mapped entry if there are multiple mappings
)

## remove NA and invalid ids
valid_indices <- !is.na(entrez_ids) & entrez_ids != ""
entrez_ids <- entrez_ids[valid_indices]
log2fc_values <- log2fc_values[valid_indices]

## combine to form a data frame
mapped_data <- data.frame(entrez_id = entrez_ids, log2fc = log2fc_values)

## resolve duplicates by selecting gene with highest log2fc value
mapped_data <- mapped_data[order(abs(mapped_data$log2fc), decreasing = TRUE), ]  # Order by absolute log2fc
mapped_data <- mapped_data[!duplicated(mapped_data$entrez_id), ]  # Remove duplicates, keeping the row with highest log2fc

## create a named vector for geneList
geneList <- mapped_data$log2fc
names(geneList) <- mapped_data$entrez_id

## sort geneList in decreasing order
geneList <- sort(geneList, decreasing = TRUE)

## check if there are still issues with geneList
if (anyDuplicated(names(geneList)) || any(is.na(names(geneList)))) {
  stop("GeneList still contains duplicate or invalid identifiers.")
}

## run gseGO
enriched_pathways3 <- gseGO(
  geneList = geneList,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP"
)

# Pool 4 markers
gene_names <- markers4$gene
log2fc_values4 <- markers4$avg_log2FC

entrez_ids4 <- mapIds(
  org.Hs.eg.db,
  keys = gene_names,
  column = "ENTREZID",
  keytype = "SYMBOL",
  multiVals = "first"  # getting the first mapped entry if there are multiple mappings
)

valid_indices4 <- !is.na(entrez_ids4) & entrez_ids4 != ""
entrez_ids4 <- entrez_ids4[valid_indices4]
log2fc_values4 <- log2fc_values4[valid_indices4]

mapped_data4 <- data.frame(entrez_id = entrez_ids4, log2fc = log2fc_values4)

mapped_data4 <- mapped_data4[order(abs(mapped_data4$log2fc), decreasing = TRUE), ]  # Order by absolute log2fc
mapped_data4 <- mapped_data4[!duplicated(mapped_data4$entrez_id), ]  # Remove duplicates, keeping the row with highest log2fc

geneList4 <- mapped_data4$log2fc
names(geneList4) <- mapped_data4$entrez_id

geneList4 <- sort(geneList4, decreasing = TRUE)

if (anyDuplicated(names(geneList4)) || any(is.na(names(geneList4)))) {
  stop("GeneList still contains duplicate or invalid identifiers.")
}

enriched_pathways4 <- gseGO(
  geneList = geneList4,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP"
)
```

```{r pathway analysis vis}
get_top_pathways <- function(enrichment_result, top_n = 10) {
  # Convert the gseGO result to a data frame
  result_df <- as.data.frame(enrichment_result)
  
  # Sort by adjusted p-value or another metric
  top_pathways <- head(result_df[order(result_df$p.adjust), ], top_n)
  
  return(top_pathways)
}

top_pathways3 <- get_top_pathways(enriched_pathways3, top_n = 10)
top_pathways4 <- get_top_pathways(enriched_pathways4, top_n = 10)

ggplot(top_pathways3, aes(x = reorder(Description, -NES), y = NES)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Most Impacted Pathways - Pool 3",
       x = "Pathway",
       y = "Normalized Enrichment Score (NES)")

ggplot(top_pathways4, aes(x = reorder(Description, -NES), y = NES)) +
  geom_bar(stat = "identity", fill = "hotpink") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 10 Most Impacted Pathways - Pool 4",
       x = "Pathway",
       y = "Normalized Enrichment Score (NES)")

top_pathways3
top_pathways4


get_top_10_enriched <- function(enriched_results) {
  result_df <- as.data.frame(enriched_results)
  # Sort by adjusted p-value
  top_10 <- result_df %>%
    arrange(p.adjust) %>%
    head(10) # Select the top 10 pathways
  # Create a new enrichment result object from the filtered data
  enriched_results@result <- enriched_results@result[row.names(enriched_results@result) %in% top_10$ID, ]
  return(enriched_results)
}

top_enriched_3 <- get_top_10_enriched(enriched_pathways3)
top_enriched_4 <- get_top_10_enriched(enriched_pathways4)

ridgeplot(top_enriched_3) + theme(axis.text.y = element_text(size = 11)) + 
  ggtitle("Top 10 Most Enriched Pathways - Pool 3")

ridgeplot(top_enriched_4) + theme(axis.text.y = element_text(size = 11)) + 
  ggtitle("Top 10 Most Enriched Pathways - Pool 4")
```
Control: The top pathways, such as "leukocyte activation" and "adaptive immune response," suggest that, in the control samples, there is a significant emphasis on immune-related pathways. This indicates that the immune system is actively engaging in normal homeostatic processes or potentially responding to regular environmental stimuli. Other top pathways like "regulation of immune system process" and "lymphocyte activation" demonstrate active regulation and initiation of immune processes in control conditions, which is expected as part of maintaining immune vigilance. The enrichment of "humoral immune response" and "immune effector process" pathways aligns with robust immune monitoring and potential response readiness, emphasizing both direct immune effector mechanisms and antibody-mediated responses.

Disease: In the disease state, the predominantly enriched pathways involve processes related to cell cycle and mitotic activities, such as "metaphase chromosome alignment" and "regulation of mitotic nuclear division." This can suggest increased cell proliferation or alterations in cell cycle regulation, common in many disease states, especially cancers. Chromosome and Nuclear Division Pathways indicate a heightened focus on cell cycle checkpoints and mitotic fidelity, potentially reflecting how the disease condition impacts cellular replication or turnover. The significant enrichment in pathways like "regulation of nuclear division" and "chromosome localization" might suggest aberrations or heightened activity in transcriptional regulation, chromosome segregation, or mitotic control, frequently observed in malignancies or proliferative disorders.






Visualize expression levels of specific genes on a dimensionality reduction plot
Visualize where specific genes are upregulated


Lets look specifically at immune cells
```{r feature plots}
# immune genes
immune_genes <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = "GO:0006955",
  keytype = "GOALL",
  columns = c("ENTREZID", "SYMBOL")
)
immune_genes <- unique(immune_genes$SYMBOL)
immune_genes3 <- immune_genes[immune_genes %in% rownames(seurat_object3)]
immune_genes4 <- immune_genes[immune_genes %in% rownames(seurat_object4)]

immune_genes3

# FeaturePlot(seurat_object3, features = immune_genes3[1:10]) 
# FeaturePlot(seurat_object4, features = immune_genes4[1:10]) 


FeaturePlot(seurat_object3, features = c("NOD2", "IL2", "IL12", "IL23", "JAK1", "TNF", "JAK3", "A4B7", "ITGA4", "CD4", "CD8"))
FeaturePlot(seurat_object4, features = c("NOD2", "IL2", "IL12", "IL23", "JAK1", "TNF", "JAK3", "A4B7", "ITGA4", "CD4", "CD8"))
```





```{r dot plot}
# Top 20
DotPlot(seurat_object3, features = unique(head(markers3$gene, 20))) + RotatedAxis()
DotPlot(seurat_object4, features = unique(head(markers4$gene, 20))) + RotatedAxis()

# Specific genes identified
DotPlot(seurat_object3, features = c("NOD2", "IL2", "IL27", "IL23", "JAK1", "TNF", "JAK3", "A4B7", "ITGA4", "CD4", "CD8")) +
  RotatedAxis() +
  ggtitle("Expression of Selected Genes in the Control Group") +
  theme(plot.title = element_text(hjust = 0.5, size = 14))

DotPlot(seurat_object4, features = c("NOD2", "IL2", "IL27", "IL23", "JAK1", "TNF", "JAK3", "A4B7", "ITGA4", "CD4", "CD8")) +
  RotatedAxis() +
  ggtitle("Expression of Selected Genes in the Disease Group") +
  theme(plot.title = element_text(hjust = 0.5, size = 14))

```
```{r selected genes}
genes <- c("NOD2", "IL2", "IL27", "JAK1", "TNF", "JAK3", "ITGA4", "CD4")

extract_gene_data <- function(seurat_obj, sample_name) {
  data <- FetchData(seurat_obj, vars = genes) %>%
    gather(key = "gene", value = "expression", genes) %>%
    group_by(gene) %>%
    summarise(average_expression = mean(expression, na.rm = TRUE)) %>%
    mutate(sample = sample_name)
  return(data)
}

control_data <- extract_gene_data(seurat_object3, "Control")
disease_data <- extract_gene_data(seurat_object4, "Disease")

combined_data <- bind_rows(control_data, disease_data) %>%
  spread(key = "sample", value = "average_expression")

combined_data %>%
  kable("html", col.names = c("Gene", "Control", "Disease")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped", position = "center")
```

```{r plot selection}
genes <- c("NOD2", "IL2", "IL27", "JAK1", "TNF", "JAK3", "ITGA4", "CD4")

extract_gene_data <- function(seurat_obj, cohort_name) {
  # Use FetchData to get the desired data
  data <- FetchData(seurat_obj, vars = genes) %>%
    gather(key = "gene", value = "expression", genes) %>%
    mutate(cohort = cohort_name)
  return(data)
}
control_data <- extract_gene_data(seurat_object3, "Control")
disease_data <- extract_gene_data(seurat_object4, "Disease")

# Combine both datasets
combined_data <- bind_rows(control_data, disease_data)

# Plot using ggplot2
ggplot(combined_data, aes(x = gene, y = expression, color = cohort)) +
  geom_jitter(aes(shape = cohort), position = position_dodge(width = 0.5), alpha = 0.6) +
  labs(title = "Expression of Selected Genes in Control and Disease Groups",
       x = "Gene", y = "Expression Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14)) +
  scale_color_manual(values = c("Control" = "blue", "Disease" = "hotpink")) +
  scale_shape_manual(values = c(16, 17))
```





predefined reference dataset HumanPrimaryCellAtlasData
```{r cell identification}
ref <- HumanPrimaryCellAtlasData()

# expression matrix
exp_matrix3 <- GetAssayData(seurat_object3, slot = "data")
exp_matrix4 <- GetAssayData(seurat_object4, slot = "data")

# singleR
cell_types3 <- SingleR(test = exp_matrix3, ref = ref, labels = ref$label.main)
cell_types4 <- SingleR(test = exp_matrix4, ref = ref, labels = ref$label.main)

# annotations
seurat_object3[['SingleR_labels']] <- cell_types3$labels
seurat_object4[['SingleR_labels']] <- cell_types4$labels

#vis
DimPlot(seurat_object3, group.by = 'SingleR_labels', reduction = "umap") +
  ggtitle("UMAP Plot for Pool 3 - Annotated with SingleR") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "UMAP 1", y = "UMAP 2")

DimPlot(seurat_object4, group.by = 'SingleR_labels', reduction = "umap") +
  ggtitle("UMAP Plot for Pool 4 - Annotated with SingleR") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "UMAP 1", y = "UMAP 2")
```


Displays expression of top marker genes across clusters
```{r heatmaps}
# Pool 3
DoHeatmap(seurat_object3, features = head(markers3$gene, 10), size = 3) +
  labs(title = "Top 10 Marker Genes Heatmap - Control") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.text.y = element_text(size = 10)
  )
# Pool 4
DoHeatmap(seurat_object4, features = head(markers4$gene, 10), size = 3) +
  labs(title = "Top 10 Marker Genes Heatmap - Disease") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.text.y = element_text(size = 10)
  )

```
